---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data



We read the file activity.csv and stored it into a df variable  

```{r readData, echo = TRUE, cache = TRUE}
df <- read.csv("activity.csv", head = TRUE)
head(df)
```

## What is mean total number of steps taken per day?  

Here we load the required packages
```{r packages, echo = FALSE, results = "hide"}
library(dplyr)
library(tidyr)

```

Now the total number of steps al calculated removing the NA values

```{r sumSteps, cache = TRUE}
df_2 <- df%>%
        group_by(date)%>%
        summarise(total_Steps= sum(steps, na.rm = TRUE))

library(knitr)
kable(df_2, format = "html")

```

A histogram of the total number of steps per day is made.
```{r }
hist(df_2$total_Steps, breaks = 10 , xlab = "Total Steps per Day", 
     main = "Frequency of Total Steps")

```

The mean and median value of the *number of steps per day* is:  

- mean =   `r mean(df_2$total_Steps)`  

- median = `r median(df_2$total_Steps)`

## What is the average daily activity pattern?

1. Series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  

```{r seriesplot, cache = TRUE}
df_3 <- df[!is.na(df$steps),]%>%
        mutate(  interval = interval*0.01)%>%
        mutate(interval = format(interval, nsmall = 2))%>%
        select(interval, steps)%>%
        group_by(interval)%>%
        summarise(mean_steps= mean(steps))
df_3$interval <- strptime(df_3$interval, "%H.%M")

with(df_3, plot(interval, mean_steps, type="l",
                main = "Average Daily Activity Pattern",
                xlab = "time interval (min.)",
                ylab = "steps average"))
```

2. Maximum value of the number of steps on daily average is `r max(df_3$mean_steps)`.
The time interval corresponding to the maximum of daily steps average is:
```{r }
x <- df_3$interval[max(df_3$mean_steps)]
x <- as.character(x)
sub("([0-9]+-[0-9]+){1,3}","",x)

```




## Imputing missing values

1. Calculating the number of NA in the dataframe:

```{r number NA}
a <- df$steps

```

number of NA: `r length(a[is.na(a)])`

2. Filling the NA values with the mean of 5-min interval calculated previously
The previous original df data frame is used
```{r fillNAsteps, cache = TRUE}
df_4 <- df[!is.na(df$steps),]%>%
        select( interval, steps)%>%
        group_by(interval)%>%
        summarise(mean_steps= mean(steps))

df_5 <- merge(df, df_4, by.x = "interval", by.y = "interval")
df_5 <- df_5%>%
        arrange(date, interval)%>%
        select(date, interval, steps, mean_steps)
for (i in 1:length(df_5$date)){
        if(is.na(df_5$steps[i])){
                df_5$steps[i]<-df_5$mean_steps[i]
        }
}


```
3. Creating a new dataset similar to the original but with NA filled in.



```{r ,cache=TRUE}
df_5 <- select(df_5, steps, date, interval)
head(df_5)
```

4. Creating a histogram as the previous one of frequency of Total Steps per Day


```{r , cache = TRUE}
df_6 <- df_5%>%
        group_by(date)%>%
        summarise(total_Steps= sum(steps, na.rm = TRUE))

library(knitr)
kable(df_6, format = "html")

```
  
  
A histogram of the total number of steps per day is made as before.

```{r, cache = TRUE}
hist(df_6$total_Steps, breaks = 10 , xlab = "Total Steps per Day", 
     main = "Frequency of Total Steps")

```

The mean and median value of the *number of steps per day* is:  

- mean =   `r mean(df_6$total_Steps)/10000` x 10^4  

- median = `r median(df_6$total_Steps)/10000` x 10^4

As it can be seen, an introduction of NA values make changes in the dataset and hence, the mean and the median value.  


## Are there differences in activity patterns between weekdays and weekends?
1. Here a new factor variable for the filled NA dataset is create. The factor distinguish between "weekday" and "weekend".
```{r , cache = TRUE}
df_7 <- df_5%>%
        mutate(type_day = weekdays(as.Date(date)))
for(i in 1:length(df_7$interval)){
        if(df_7$type_day[i]=="sábado" |df_7$type_day[i]=="domingo" ){
                df_7$type_day[i]<-"weekend"
        }
        else{df_7$type_day[i] <- "weekday"}
}
head(df_7,10)
```

2. Creating a dataframe for plotting the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r cache = TRUE}
df_7 <- df_7%>%
        group_by(interval, type_day)%>%
        summarise(mean_steps = mean(steps))%>%
        mutate(type_day = as.factor(type_day))%>%
        arrange(mean_steps, type_day)

```
2. Creating a panel plot 
```{r}
library(lattice)
xyplot(mean_steps ~ interval  | type_day, data = df_7, layout = c(1,2),
       xlab = "Interval", ylab= "Number of steps", type="l")      

```
